
First and foremost, I would like to thank my advisor Dr. Bruno C. d. S. Oliveira
for his continued support and mentorship throughout my studies. It is my honor
to be one of his first disciples. Back in 2014 when I was in my final year at
ZJU, I wasn't sure if I wanted to pursue a PhD, let alone a PhD in Programming
Languages! My fellow classmates were all considering applying for PhD in AI,
computer vision, etc; there were no courses on the theories of programming
languages in the computer science curriculum in most (if not all) universities
in Mainland China. My first encounter with this field was due to a Coursera
course I happened to enroll in. Immediately I got fascinated and wanted to dig
more. Looking back, it was such a great timing that Bruno just landed his
faculty position as an assistant professor at HKU, and that I just finished my
undergraduate studies. Our initial conversation went very well, and shortly I
became his first PhD student. For the first few months, I was quite ``on the
right track''. I recall a conversation with my colleague (Haoyuan, is that you?)
that Bruno referred me as the ``push model'' because I always brought lots of
questions in our weekly meetings and pushed him to give me more tasks. It was
not long before I lost myself, had difficulty making any progress, and that the
weekly meetings with Bruno soon became nightmares to me. Bruno, thank you for
not giving me up, encouraging me to keep reading more papers, keep trying
different ideas no matter how hopeless they may seem, and especially for not
reminding me of the possibility that I may not be able to finish PhD on time!
All in all, I could not wish for a better PhD advisor.

I am incredibly fortunate to have met Prof. Tom Schrijvers from KU Leuven,
talked to him about my work and found a topic of mutual interest, which led to
one of the key publications for this PhD thesis. I learned a lot from you! I
would also like to thank my co-advisor Prof. T.H. Tse for his valuable
suggestions and guidelines. He helped revise my papers and this thesis,
continuously nudging me to make my writing clearer. His feedback is nothing
short of inspiring, which I greatly enjoyed. During the years I spent at HKU, I
had the opportunity to collaborate with a number of excellent fellow
researchers, to name a few (in no particular order): Tomas Tauber, Zhiyuan Shi,
Weixin Zhang, Huang Li, Yanpeng Yang, Ningning Xie, Yanlin Wang and Haoyuan
Zhang. Especially my coauthor Ningning Xie, who, during my low days, offered me
to collaborate on a project of her. This collaboration kept me busy for a while, occupying the time I would otherwise spend on indulging in the gloomy
outlook of my research life. Our fruitful collaboration is still going on as of
this writing.

Last but not least, I would love to thank my family for being extremely
supportive throughout my studies. I made it through thanks to all of you.