%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{First-Class Traits}
\label{chap:traits}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\renewcommand\ottaltinferrule[4]{
  \inferrule*[narrower=0.9,lab=#1,#2]
    {#3}
    {#4}
}

In this chapter and \cref{chap:case_study}, we present two applications of
\fnamee. This chapter is concerned with building a source-level language named
\sedel that features \textit{typed first-class traits}. In particular we show
how to model source-level constructs for first-class traits and dynamic
inheritance, supporting standard object-oriented features such as dynamic
dispatching and abstract methods. It is remarkable that all of these can be
explained by plain \fnamee expressions, showing its expressive power.
In \cref{chap:case_study} we conduct a case study of modularizing programming
language features by the means of first-class traits.



\section{Motivation: First-Class Classes and Dynamic Inheritance}

Many dynamically typed-languages (including JavaScript, Ruby, Python
or Racket) support \emph{first-class classes}~\citep{DBLP:conf/aplas/FlattFF06}, or related concepts
such as first-class mixins and/or traits. In those languages classes
are first-class values and, like any other values, they can be
passed as an argument, or returned from a function. Furthermore
first-class classes support \emph{dynamic inheritance}: i.e., they
can inherit from other classes at \emph{run time}, enabling
programmers to abstract over the inheritance hierarchy.
Those features make first-class classes very powerful and expressive,
and enable highly modular and reusable pieces of code, such as:
\begin{lstlisting}[language=JavaScript]
const mixin = Base => {
  return class extends Base { ... }
};
\end{lstlisting}
In this piece of JavaScript code, \lstinline{mixin} is
parameterized by a class \lstinline{Base}. Note that the concrete
implementation of \lstinline{Base} can be
even dynamically determined at run time, for example
after reading a configuration file to decide which
class to use as the base class.  When applied to an argument,
\lstinline{mixin} will create a new class on-the-fly and return that
as a result. Later that class can be instantiated and used to create
new objects, as any other classes.

In contrast, most statically-typed
languages do not have first-class classes and dynamic
inheritance. While all statically-typed OO languages allow first-class
\emph{objects} (i.e., objects can be passed as arguments and returned
as results), the same is not true for classes. Classes in languages such as
Scala, Java or C++ are typically a second-class construct, and the
inheritance hierarchy is \emph{statically determined}. The closest thing
to first-class classes in
languages like Java or Scala are classes such as
\lstinline[language=java]{java.lang.Class} that enable representing classes and
interfaces as part of their reflective framework. \lstinline[language=java]{java.lang.Class} can be used to
mimic some of the uses of first-class classes, but in an essentially
dynamically-typed way. Furthermore simulating first-class classes
using such mechanisms is highly cumbersome because classes need to be
manipulated programmatically. For example instantiating a new class
cannot be done using the standard \lstinline{new} construct, but
rather requires going through API methods of
\lstinline[language=java]{java.lang.Class}, such as \lstinline{newInstance}, for
creating a new instance of a class.

Despite the popularity and expressive power of first-class classes in dynamically-typed
languages, there is surprisingly little work on typing of first-class
classes (or related concepts such as first-class mixins or traits).
First-class classes and dynamic inheritance pose well-known
difficulties in terms of typing. For example, in his thesis,
\citet{bracha1992programming} comments several times on the difficulties of typing
dynamic inheritance and first-class mixins, and proposes the
restriction to static inheritance that is also common in
statically-typed languages. He also observes that such restriction
poses severe limitations in terms of expressiveness, but that appeared
(at the time)
to be a necessary compromise when typing was also desired.
Only recently some progress has been made in statically typing
first-class classes and dynamic inheritance. In particular there are
two works in this area: Racket's gradually
typed first-class classes~\citep{DBLP:conf/oopsla/TakikawaSDTF12}; and \citeauthor{DBLP:conf/ecoop/LeeASP15}'s model of
typed first-class classes~\citep{DBLP:conf/ecoop/LeeASP15}. Both works provide typed models of
first-class classes, and they enable encodings of mixins~\citep{bracha1990mixin}
similar to those employed in dynamically-typed languages.

However, as far as we known no previous work supports statically-typed
\emph{first-class traits}. Traits~\citep{scharli2003traits, Ducasse_2006} are an
alternative to mixins, and other models of (multiple) inheritance. The key
difference between traits and mixins lies on the treatment of conflicts when
composing multiple traits/mixins. Mixins adopt an \emph{implicit} resolution
strategy for conflicts, where the compiler automatically picks one
implementation in case of conflicts. For example, Scala uses the order of mixin
composition to determine which implementation to pick in case of conflicts.
Traits, on the other hand, employ an \emph{explicit} resolution strategy, where
the compositions with conflicts are rejected, and the conflicts are explicitly
resolved by programmers. This gives programmers fine-grained control, when
conflicts arise, of selecting desired features from different components. Thus
we believe traits are a better model for multiple inheritance in
statically-typed object-oriented languages. In what follows, we present \sedel:
the first design of typed first-class traits.



\input{Sources/Traits/overview.tex}
\input{Sources/Traits/traits.tex}
\input{Sources/Traits/typesystem.mng}
% \input{Sources/Traits/applications.tex}
% \input{Sources/Traits/conclusion.tex}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../Thesis"
%%% org-ref-default-bibliography: ../Thesis.bib
%%% End:
